Predicting the effects of character displacement on food-web dynamics

### 2019-02-09 manuscript_ECD_model.Rmd
add Borrelli cites
edits to in text citations

Ecological character displacement is thought to be a key evolutionary process generating biodiversity [@Schluter2000; @Pfennig2010; but see @Stuart2013]. Ecological character displacement describes the “process of phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has been generated to understand the scenarios under which exploitative competition for resources leads to the divergence of consumer foraging traits. The general result emerging from these studies is that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the divergence of consumer foraging traits [@Lawlor1976; @Taper1985). This process is not simply driven by ecological differences, but creates an eco-evolutionary feedback that drives further differentiation. This key insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

@@

Although explicit models of resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of competition and coexistence theory [Lawlor1976; Germain2018; Bassar; McPeek]. 
For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. 
<!--
On the other hand, specialization often increases a consumer’s ability to capture a specific resource.
From food-web theory, we know that increased feeding rates often lead to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics [Rosenzweig 1971; Murdoch et al. 2003; McCann 2011].
-->
Yet, this reduction in interspecific competition may, at the same time, increase the interspecific interaction between a consumer and the resource it specializes on.  
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often leads to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

@@

Here, I address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these ecological effects, I examined three different foraging scenarios: *CONDENSE THIS. NO NEED TO GIVE ALL THE DETAILS YET. RIGHT NOW IT IS JUST CONFUSING.* (i) resources occur in the same habitat and consumers exhibit a linear functional response (e.g. @MacArthur1972; @Lawlor1976); (ii) resources occur in different habitats and consumers exhibit a linear functional response (e.g. @Lawlor1976); (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I also explore the contingency of these effects to different assumptions about evolutionary trade-offs in consumer foraging traits. I focus on the ecological side of these dynamics because the evolutionary aspects have been well characterized [@Lawlor1976; @Taper1985; @Abrams1986; McPeekRECENT]. I show that the consequences of character displacement on food-web dynamics are not always intuitive and depend on the foraging scenario and tradeoff in consumer traits.

@@

<!-- Food-web stability -->
Our most stiking result was that character displacement, when examined through the lens of food-web theory, usually resulted in less stable food webs. 
In fact, under a foraging scenario that incorporates more biological realism, character displacement may even result in an unstable food web. 
This result aligns well with the general result from food-web theory that food-web stability tends to go hand-in-hand with resource densities [@May1973; @Murdoch2003; @McCann2011]. 
However, this result contrasts the notion from competition theory that character displacement contributes to species coexistence. 
My results do not contradict these past inferences, as they were based on a different measure of stability that considers the ability of a consumer with a different phenotype to invade when rare (invasion criteria). 
However, its clear that stability measures do not always go hand-in-hand and that a pluralistic view of different aspects of stability is needed [@Donohue2016] for understanding eco-evolutionary feedbacks on food-web dynamics.

<!-- Ghost of competition and stability past --->
A single species of threespine stickleback lives in hundreds of lakes of British Columbia, but the species pairs, where character displacement has happened, are only known from four lakes one of which has undergone reverse speciation [@Rudman2016]. 
Our results hint at an interesting hypothesis for the apparent rarity, namely founded on ecological stability [@Borrelli2015]. 
Specifically, the commonness of the lakes with only a single species is due to the higher ecological stability of this food-web structure, compared to the one with a species pair. 
Testing this hypothesis may be quite difficult, but it could be possible with recent advances in genomics. 
For example, the species-pair lake that has undergone reverse speciation has a different ecological impact than other generalist lakes in that it is much more like the benthic-ecotype in its morphology. It is unclear whether this is a transient or persistent evolutionary state. 
But advances in genomics may be able to tease apart whether a currently generalist species had a genomic imprint of being more benthic or limnetic.

@@

<!-- Caveats -->
Although I used an Adaptive Dynamics approach here, I expect that these theoretical predictions would apply to models that assumed a different genetic architecture of attack rates (e.g. quantitative genetics). 
I expect this because, time and time again, models that explicitly include resource dynamics inevitably show that resource competition results in divergent character displacement, regardless of the genetic architecture of traits (see synthesis by @Taper1985). 
Since these models are focused primarily on trait evolution, it is not clear that divergence in these models would have different ecological effects.
In fact, recent work by @McPeek2019 has shown that genetic divergence in a quantitative genetics model can lead to decreased resource abundances and unstable dynamics.
Finally, my conclusions may only apply to biological resources that are nutritionally substitutable. 
It would be interesting to extend these current analyses to non-substitutable resources where convergent character displacement is expected (@Abrams1987; @Fox2008).

>>>>

Ecological character displacement is thought to be a key evolutionary process generating biodiversity [@Schluter2000; @Pfennig2010; but see @Stuart2013]. Ecological character displacement describes the “process of phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has been generated to understand the scenarios under which exploitative competition for resources leads to the divergence of consumer foraging traits. The general result emerging from these studies is that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the divergence of consumer foraging traits [@Lawlor1976; @Taper1985]. This process is not simply driven by ecological differences, but creates an eco-evolutionary feedback that drives further differentiation. This key insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

@@

Although explicit models of resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of competition and coexistence theory [@Lawlor1976; @Germain2018; @Bassar; @McPeek2019]. 
For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. 
<!--
On the other hand, specialization often increases a consumer’s ability to capture a specific resource.
From food-web theory, we know that increased feeding rates often lead to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics [Rosenzweig 1971; Murdoch et al. 2003; McCann 2011].
-->
Yet, this reduction in interspecific competition may, at the same time, increase the interspecific interaction between a consumer and the resource it specializes on.  
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often leads to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

@@

Here, I address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these ecological effects, I examined three different foraging scenarios: *CONDENSE THIS. NO NEED TO GIVE ALL THE DETAILS YET. RIGHT NOW IT IS JUST CONFUSING.* (i) resources occur in the same habitat and consumers exhibit a linear functional response [e.g. @MacArthur1972; @Lawlor1976]; (ii) resources occur in different habitats and consumers exhibit a linear functional response [e.g. @Lawlor1976]; (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I also explore the contingency of these effects to different assumptions about evolutionary trade-offs in consumer foraging traits. I focus on the ecological side of these dynamics because the evolutionary aspects have been well characterized [@Lawlor1976; @Taper1985; @Abrams1986; McPeekRECENT]. I show that the consequences of character displacement on food-web dynamics are not always intuitive and depend on the foraging scenario and tradeoff in consumer traits.

@@

<!-- Food-web stability -->
Our most stiking result was that character displacement, when examined through the lens of food-web theory, usually resulted in less stable food webs. 
In fact, under a foraging scenario that incorporates more biological realism, character displacement may even result in an unstable food web. 
This result aligns well with the general result from food-web theory that food-web stability tends to go hand-in-hand with resource densities [@May1973; @Murdoch2003; @McCann2011]. 
However, this result contrasts the notion from competition theory that character displacement contributes to species coexistence. 
My results do not contradict these past inferences, as they were based on a different measure of stability that considers the ability of a consumer with a different phenotype to invade when rare (invasion criteria). 
However, its clear that stability measures do not always go hand-in-hand and that a pluralistic view of different aspects of stability is needed [@Donohue2013; @Donohue2016] for understanding eco-evolutionary feedbacks on food-web dynamics.

<!-- Ghost of competition and stability past --->
A single species of threespine stickleback lives in hundreds of lakes of British Columbia, but the species pairs, where character displacement has happened, are only known from four lakes one of which has undergone reverse speciation [@Rudman2016]. 
Our results hint at an interesting hypothesis for the apparent rarity, namely founded on ecological stability [@Borrelli2015a; @Borrelli2015b]. 
Specifically, the commonness of the lakes with only a single species is due to the higher ecological stability of this food-web structure, compared to the one with a species pair. 
Testing this hypothesis may be quite difficult, but it could be possible with recent advances in genomics. 
For example, the species-pair lake that has undergone reverse speciation has a different ecological impact than other generalist lakes in that it is much more like the benthic-ecotype in its morphology. It is unclear whether this is a transient or persistent evolutionary state. 
But advances in genomics may be able to tease apart whether a currently generalist species had a genomic imprint of being more benthic or limnetic.

@@

<!-- Caveats -->
Although I used an Adaptive Dynamics approach here, I expect that these theoretical predictions would apply to models that assumed a different genetic architecture of attack rates (e.g. quantitative genetics). 
I expect this because, time and time again, models that explicitly include resource dynamics inevitably show that resource competition results in divergent character displacement, regardless of the genetic architecture of traits [see synthesis by @Taper1985]. 
Since these models are focused primarily on trait evolution, it is not clear that divergence in these models would have different ecological effects.
In fact, recent work by @McPeek2019 has shown that genetic divergence in a quantitative genetics model can lead to decreased resource abundances and unstable dynamics.
Finally, my conclusions may only apply to biological resources that are nutritionally substitutable. 
It would be interesting to extend these current analyses to non-substitutable resources where convergent character displacement is expected [@Abrams1987; @Fox2008].
